An elaborate data set on human gait and the effect of mechanical perturbations

### 2014-12-15 paper.tex
Preparation for PeerJ preprint submission.
- Added methods section.
- Changed to 2.5cm margins.
- Changed to 12pt font.
- Repaired some citations.
- Added line numbers.

\documentclass[fleqn,10pt]{wlpeerj}

@@

\section*{Introduction}
%
The collection of dynamical data during human walking has a long history
beginning with the first motion pictures and now with modern marker based
motion capture techniques and high fidelity ground reaction load measurements.
Even though years of data on thousands of subjects now exist, this data is not
widely disseminated, well organized, nor available with few or no restrictions.
David Winter's published normative gait data~\cite{Winter1990} is widely used
in biomechanical studies, yet it comes from relatively few subjects and only a
small number of gait cycles per subject. This small source has successfully
inspired many other studies, such as powered prosthetic control
design~\cite{Sup2008}, but success in other research fields using large sets of
data for discovery lead one to believe that more elaborate data sets may
benefit the field of human motion studies. To enable such work, biomechanical
data needs to be shared extensively, organized, and curated to enabled
future analysts.

There are some notable gait data sets and databases besides Winter's
authoritative set that are publicly available. The International Society of
Biomechanics has maintained a web page (http://isbweb.org/data) since
approximately 1995 that includes data sets for download and mostly unencumbered
use. For example, Vaughn, et. al's data~\cite{Vaughan1992} with kinematics and
force plate measurement from several subjects is available on the site. At
another website, the CGA Normative Gait Database~\cite{Kirtley2014}, Chris
Kirtley shares normative gait data from several studies and these files have
influenced other studies, for example the average gait cycles from children
used in~\cite{Bogert2003}.

Chester et. al~\cite{Chester2007} report on a large gait database comparison
where one database contained kinematic data of 409 gait cycles of children from
1 to 7 years old but the data does not seem to be publicly available. This is
unfortunately typical. But Tirosh et. al, recognized the need for a
comprehensive data base for clinical gait data and created the
Gaitabase~\cite{Tirosh2010}. This database may contain a substantial amount of
data but it is encumbered by a very complicated and restrictive license and
sharing scheme. However, there are examples of data with less restrictions. The
University of Wisconsin at LaCrosse has an easily accessible normative gait
data set~\cite{Wilson2014} from 25 subjects with lower extremity marker data
from multiple gait cycles and force plate measurements from a single gait
cycle.

@@

With all of this in mind, we collected over seven and half hours of gait data
from fifteen able bodied subjects which amounts to over 25,000 gait cycles. The
subjects walked at three different speeds on an instrumented treadmill while we
collected full body marker locations and ground reaction loads from a pair of
force plates. The protocol for the majority of the trials included two minutes
of normal walking and eight minutes of walking under the influence of
pseudo-random belt speed fluctuations. The data has been organized complete
with rich meta data and made available in the most unrestrictive form for other
research uses following modern best practices in data sharing~\cite{White2013}.

@@

\section*{Participants}
%
Fifteen able bodied subjects including four females and eleven males with an
average age of $24\pm4$ years, height of $1.75\pm0.09$ m, mass of $74\pm13$ kg
participated in the study. The study was approved by the Institutional Review
Board of Cleveland State University (\# 29904-VAN-HS) and informed consent was
obtained from all participants. The data has been anonymized with respect to the
participants' identities and a unique identification number was assigned to
each subject. A selection of the meta data collected for each subject is shown
in Table~\ref{tab:subjects}.
%
\begin{table*}
  \cprotect\caption{Information about the 15 participants. The final three
    columns give the trial numbers associated with each nominal treadmill
    speed. The measured mass is computed from the mean total vertical ground
    reaction force just after the calibration pose event, if possible.
    Additional trials found in the data set with a subject identification
    number 0 are trials with no subject, i.e.  unloaded trials that can be used
    for inertial compensation purposes, and are not shown in the table. Generated
    by \verb|src/subject_table.py|.}
  \centering
  \small
  \input{tables/subjects.tex}
  \label{tab:subjects}
\end{table*}

\section*{Equipment}
%
The data were collected in the Laboratory for Human Motion and Control at
Cleveland State University, using the following equipment:

\begin{itemize}
  \item A R-Mill treadmill which has dual 6 degree of freedom force plates,
    independent belts for each foot, and lateral/pitch motion capabilities
    (Forcelink, Culemborg, Netherlands).
  \item A 10 Osprey camera motion capture system paired with the Cortex
    3.1.1.1290 software (Motion Analysis, Santa Rosa, CA, USA).
  \item USB-6255 data acquisition unit (National Instruments, Austin, Texas,
    USA).
  \item Four ADXL330 Triple Axis Accelerometer Breakout boards attached to the
    treadmill (Sparkfun, Niwot, Colorado, USA).
  \item D-Flow software (versions 3.16.1 to 3.16.2) and visual display system,
    (Motek Medical, Amsterdam, Netherlands).
\end{itemize}

@@

\section*{Protocol}
%
The experimental protocol consisted of both static measurements and walking on
the treadmill for 10 minutes under unperturbed and perturbed conditions. Before
a set of trials on the same day the following happened:
%
\begin{itemize}
  \item Calibration of the motion capture system using the manufacturer's
    recommended procedure.
  \item Subject changes into athletic shoes, shorts, sports bra, baseball
    cap, and rock climbing harness.
  \item All 47 markers are applied directly to the skin except for the heel,
    toe, and head markers, which were placed on the respective article of
    clothing.~\footnote{The sacrum and rear pelvic markers may have been placed
    on the shorts for a small number of the subjects}.
  \item Subjects self-reported age, gender, and mass.
  \item Height was measured by the experimentalist.
  \item Four reference photographs (front, back, right, left) were taken of
    subject's marker locations.
\end{itemize}

@@

\section*{Perturbation Signals}
%
As previously described, the protocol included a phase of normal walking,
followed by longitudinal belt speed perturbations, and ended with a second
segment of normal walking. Three pseudo-random belt speed control signals, with
mean velocities of \SIlist{0.8;1.2;1.6}{\meter\per\second}, were pre-generated
with MATLAB and Simulink (Mathworks, Natick, Massachusetts, USA).  The same
control signal was used for all trials at that given speed.

@@

To show the effects of the treadmill dynamics and give an idea of the frequency
content of the actual perturbations, the input and output for each speed were
transformed into the frequency domain using the Fast Fourier algorithm, and the
results are shown in Figure~\ref{fig:freq_analysis}. This shows that for the
1.2~\si{\meter\per\second} walking speed, the amplitude of the output is
significantly lower than the amplitude of the input signal at lower
frequencies. Additionally, the amplitude of output signal in the
0.8~\si{\meter\per\second}  walking speed begins to attenuate around
2~\si{\hertz}, which is a noticeably lower frequency than the other walking
speeds.
%
\begin{figure}
  \centering
  \includegraphics{figures/frequency_analysis.pdf}
  \caption{Frequency spectrum of the treadmill belt velocity input signal
    (purple) and the recorded output velocity (blue) for average belt speeds of
    0.8, 1.2, and 1.6 m/s, respectively}
  \label{fig:freq_analysis}
\end{figure}

\section*{Raw Data}
%
The raw data consists of a set of ASCII tab delimited text files output from
both the ``mocap'' and ``record'' modules in D-Flow in addition to a manually
generated YAML file that contains all of the necessary meta data for the given
trial. These three files are stored in a hierarchy of directories with one
trial per directory. The directories are named in the following fashion
\verb+T001/+ where \verb+T+ stands for ``trial'' and the following three digits
are provide a unique trial identification number.

\subsection*{mocap-xxx.txt}

@@

\subsection*{record-xxx.txt}
%
The record module also outputs a tab delimited ASCII text file with numerical
values at six decimal digits. It includes a \verb+Time+ column which records
the D-Flow system time in seconds. This time corresponds to the time recorded
in the \verb+TimeStamp+ column in mocap module tsv file which is necessary for
time synchronization. There are two additional columns \verb+RightBeltSpeed+
and \verb+LeftBeltSpeed+ which provide the independent belt speeds measured in
meters per second by a factory installed encoder in the treadmill.

@@

\subsection*{meta-xxx.yml}

@@

\section*{Processed Data}
%
We developed a toolkit for data processing, GaitAnalysisToolKit v0.1.2,
\cite{Moore2014a}, for common gait computations and provide an example
processed trial to present the nature of the data. The tool was developed in
Python, is dependent on the SciPy Stack and Octave, and provides two main
classes: one to do basic gait data cleaning from D-Flow's output files,
\verb|DFlowData|, and a second to compute common gait variables of interest,
\verb|GaitData|.

@@

\section*{Data Limitations}
%
The data is provided in good faith with great attention to detail but as with
all data there are anomalies that may affect the use and interpretation of
results emanating from the data. The following list gives various notes and
warnings about the data that should be taken into account when making use of
it.
%
\begin{itemize}
  \item Be sure to read the notes in each meta data file for details about
    possible anomalies in that particular trial. Things such as marker dropout,
    ghost markers, and marker movement are the more prominent notes. Details
    about variations in the equipment on the day of the trial are also
    mentioned.
  \item The subject identification number 0 stands for "no subject" and was
    used whenever data was collected from the system with no subject on the
    treadmill, for example during the trials that were intended to be used for
    inertial compensation purposes. These trials play through the exact
    protocol as those with a human subject and the matching trials are
    indicated in the meta data. Matching unloaded trials were recorded on the
    same day as the loaded trials and is noted in the
    \verb|trial:files:compensation| section of the meta data file.
  \item Trials 1 and 2 were not recorded as part of this study. Those trial
    identification numbers were reserved for early data exploration from data
    collected in other studies.
  \item Trials 37, 38, and 39 do not exist. The numbers were accidentally
    skipped.
  \item Trials 9, 10, and 11 used a slightly different event definition where the
    calibration poses were not explicitly tagged by an event, yet the protocol
    was the identical to the following trials. The calibration pose will have
    to be determined manually.
  \item  Trials 6-15 have force measurements are affected by the treadmill
    vibration mode mentioned in the equipment section and the forces should be
    not be used. We include the trials because both the kinematic data is valid
    and trials 6-8 include lateral perturbations in addition to the
    longitudinal.
  \item During trials 9-15 we used wooden blocks to fix the treadmill to the
    concrete floor to eliminate the treadmill's low vibration mode
    (\~5\si{\hertz}). But these blocks seem to have corrupted the force plate
    measurements by imposing frictional stresses on the system. The force plate
    measurements should not be used from these trials, but the marker data is
    fine.
  \item Trials 6-8 use an early experimental protocol which divided the
    perturbation sections into three sections: longitudinal perturbations,
    lateral perturbations, and a combination of each. We then learned the
    treadmill had a low vibrational mode which significantly affects the force
    plate measurements, requiring us to eliminate the lateral perturbation
    motions. The force measurements during these trials are corrupted by this
    vibrational mode and should be used with caution or not at all.
  \item We did not record unloaded compensation trials for trials 9-15.
    Regardless, they would likely be useless due to the corruption from the
    wooden blocks.
  \item Trials 6-8 use a only the lower body marker set. The remaining trials
    are full body.
  \item The ankle joint torques computed from subject 9's data in trials 25-27
    are abnormal and should be used with caution or not at all. We were not
    able to locate the source of the error, but it is likely related to the
    force calibration.
\end{itemize}

>>>>

% 12pt font is recommend for the preprint
\documentclass[fleqn,12pt]{wlpeerj}

% for the preprint we add line numbers
\usepackage{lineno}
\linenumbers

@@

\section*{Introduction}
%
The collection of dynamical data during human walking has a long history
beginning with the first motion pictures and now with modern marker based
motion capture techniques and high fidelity ground reaction load measurements.
Even though years of data on thousands of subjects now exist, this data is not
widely disseminated, well organized, nor available with few or no restrictions.
David Winter's published normative gait data, \cite{Winter1990}, is widely used
in biomechanical studies, yet it comes from relatively few subjects and only a
small number of gait cycles per subject. This small source has successfully
inspired many other studies, such as powered prosthetic control
design, \cite{Sup2008}, but success in other research fields using large sets
of data for discovery lead one to believe that more elaborate data sets may
benefit the field of human motion studies. To enable such work, biomechanical
data needs to be shared extensively, organized, and curated to enabled future
analysts.

There are some notable gait data sets and databases besides Winter's
authoritative set that are publicly available. The International Society of
Biomechanics has maintained a web page (http://isbweb.org/data) since
approximately 1995 that includes data sets for download and mostly unencumbered
use. For example, Vaughn, et. al's data, \cite{Vaughan1992}, with kinematics
and force plate measurement from several subjects is available on the site. At
another website, the CGA Normative Gait Database, \cite{Kirtley2014}, Chris
Kirtley shares normative gait data from several studies and these files have
influenced other studies, for example the average gait cycles from children
used in \cite{Bogert2003}.

Chester et. al, \cite{Chester2007}, report on a large gait database comparison
where one database contained kinematic data of 409 gait cycles of children from
1 to 7 years old but the data does not seem to be publicly available. This is
unfortunately typical. But Tirosh et. al, recognized the need for a
comprehensive data base for clinical gait data and created the
Gaitabase, \cite{Tirosh2010}. This database may contain a substantial amount of
data but it is encumbered by a very complicated and restrictive license and
sharing scheme. However, there are examples of data with less restrictions. The
University of Wisconsin at LaCrosse has an easily accessible normative gait
data set, \cite{Willson2014}, from 25 subjects with lower extremity marker data
from multiple gait cycles and force plate measurements from a single gait
cycle.

@@

With all of this in mind, we collected over seven and half hours of gait data
from fifteen able bodied subjects which amounts to over 25,000 gait cycles. The
subjects walked at three different speeds on an instrumented treadmill while we
collected full body marker locations and ground reaction loads from a pair of
force plates. The protocol for the majority of the trials included two minutes
of normal walking and eight minutes of walking under the influence of
pseudo-random belt speed fluctuations. The data has been organized complete
with rich meta data and made available in the most unrestrictive form for other
research uses following modern best practices in data sharing, \cite{White2013}.

@@

\section*{Methods}
\subsection*{Participants}
%
Fifteen able bodied subjects including four females and eleven males with an
average age of $24\pm4$ years, height of $1.75\pm0.09$ m, mass of $74\pm13$ kg
participated in the study. The study was approved by the Institutional Review
Board of Cleveland State University (\# 29904-VAN-HS) and informed consent was
obtained from all participants. The data has been anonymized with respect to the
participants' identities and a unique identification number was assigned to
each subject. A selection of the meta data collected for each subject is shown
in Table~\ref{tab:subjects}.
%
\begin{table*}
  \cprotect\caption{Information about the 15 participants. The final three
    columns give the trial numbers associated with each nominal treadmill
    speed. The measured mass is computed from the mean total vertical ground
    reaction force just after the calibration pose event, if possible.
    Additional trials found in the data set with a subject identification
    number 0 are trials with no subject, i.e.  unloaded trials that can be used
    for inertial compensation purposes, and are not shown in the table. Generated
    by \verb|src/subject_table.py|.}
  \centering
  \small
  \input{tables/subjects.tex}
  \label{tab:subjects}
\end{table*}

\subsection*{Equipment}
%
The data were collected in the Laboratory for Human Motion and Control at
Cleveland State University, using the following equipment:
%
\begin{itemize}
  \item A R-Mill treadmill which has dual 6 degree of freedom force plates,
    independent belts for each foot, and lateral/pitch motion capabilities
    (Forcelink, Culemborg, Netherlands).
  \item A 10 Osprey camera motion capture system paired with the Cortex
    3.1.1.1290 software (Motion Analysis, Santa Rosa, CA, USA).
  \item USB-6255 data acquisition unit (National Instruments, Austin, Texas,
    USA).
  \item Four ADXL330 Triple Axis Accelerometer Breakout boards attached to the
    treadmill (Sparkfun, Niwot, Colorado, USA).
  \item D-Flow software (versions 3.16.1 to 3.16.2) and visual display system,
    (Motek Medical, Amsterdam, Netherlands).
\end{itemize}

@@

\subsection*{Protocol}
%
The experimental protocol consisted of both static measurements and walking on
the treadmill for 10 minutes under unperturbed and perturbed conditions. Before
a set of trials on the same day the following happened:
%
\begin{itemize}
  \item Calibration of the motion capture system using the manufacturer's
    recommended procedure.
  \item Subject changes into athletic shoes, shorts, sports bra, baseball
    cap, and rock climbing harness.
  \item All 47 markers are applied directly to the skin except for the heel,
    toe, and head markers, which were placed on the respective article of
    clothing.~\footnote{The sacrum and rear pelvic markers may have been placed
    on the shorts for a small number of the subjects}.
  \item Subjects self-reported age, gender, and mass.
  \item Height was measured by the experimentalist.
  \item Four reference photographs (front, back, right, left) were taken of
    subject's marker locations.
\end{itemize}

@@

\subsection*{Perturbation Signals}
%
As previously described, the protocol included a phase of normal walking,
followed by longitudinal belt speed perturbations, and ended with a second
segment of normal walking. Three pseudo-random belt speed control signals, with
mean velocities of \SIlist{0.8;1.2;1.6}{\meter\per\second}, were pre-generated
with MATLAB and Simulink (Mathworks, Natick, Massachusetts, USA).  The same
control signal was used for all trials at that given speed.

@@

To show the effects of the treadmill dynamics and give an idea of the frequency
content of the actual perturbations, the input and output for each speed were
transformed into the frequency domain using the Fast Fourier algorithm, and the
results are shown in Figure~\ref{fig:freq_analysis}. This shows that for the
1.2~\si{\meter\per\second} walking speed, the amplitude of the output is
significantly lower than the amplitude of the input signal at lower
frequencies. Additionally, the amplitude of output signal in the
0.8~\si{\meter\per\second}  walking speed begins to attenuate around
2~\si{\hertz}, which is a noticeably lower frequency than the other walking
speeds.
%
\begin{figure}
  \centering
  \includegraphics{figures/frequency_analysis.pdf}
  \caption{Frequency spectrum of the treadmill belt velocity input signal
    (purple) and the recorded output velocity (blue) for average belt speeds of
    0.8, 1.2, and 1.6 m/s, respectively}
  \label{fig:freq_analysis}
\end{figure}
\section*{Results}
\subsection*{Raw Data}
%
The raw data consists of a set of ASCII tab delimited text files output from
both the ``mocap'' and ``record'' modules in D-Flow in addition to a manually
generated YAML file that contains all of the necessary meta data for the given
trial. These three files are stored in a hierarchy of directories with one
trial per directory. The directories are named in the following fashion
\verb+T001/+ where \verb+T+ stands for ``trial'' and the following three digits
are provide a unique trial identification number.

\subsubsection*{mocap-xxx.txt}

@@

\subsubsection*{record-xxx.txt}
%
The record module also outputs a tab delimited ASCII text file with numerical
values at six decimal digits. It includes a \verb+Time+ column which records
the D-Flow system time in seconds. This time corresponds to the time recorded
in the \verb+TimeStamp+ column in mocap module tsv file which is necessary for
time synchronization. There are two additional columns \verb+RightBeltSpeed+
and \verb+LeftBeltSpeed+ which provide the independent belt speeds measured in
meters per second by a factory installed encoder in the treadmill.

@@

\subsubsection*{meta-xxx.yml}

@@

\subsection*{Processed Data}
%
We developed a toolkit for data processing, GaitAnalysisToolKit v0.1.2,
\cite{Moore2014a}, for common gait computations and provide an example
processed trial to present the nature of the data. The tool was developed in
Python, is dependent on the SciPy Stack and Octave, and provides two main
classes: one to do basic gait data cleaning from D-Flow's output files,
\verb|DFlowData|, and a second to compute common gait variables of interest,
\verb|GaitData|.

@@

\subsection*{Data Limitations}
%
The data is provided in good faith with great attention to detail but as with
all data there are anomalies that may affect the use and interpretation of
results emanating from the data. The following list gives various notes and
warnings about the data that should be taken into account when making use of
it.
%
\begin{itemize}
  \item Be sure to read the notes in each meta data file for details about
    possible anomalies in that particular trial. Things such as marker dropout,
    ghost markers, and marker movement are the more prominent notes. Details
    about variations in the equipment on the day of the trial are also
    mentioned.
  \item The subject identification number 0 stands for "no subject" and was
    used whenever data was collected from the system with no subject on the
    treadmill, for example during the trials that were intended to be used for
    inertial compensation purposes. These trials play through the exact
    protocol as those with a human subject and the matching trials are
    indicated in the meta data. Matching unloaded trials were recorded on the
    same day as the loaded trials and is noted in the
    \verb|trial:files:compensation| section of the meta data file.
  \item Trials 1 and 2 were not recorded as part of this study. Those trial
    identification numbers were reserved for early data exploration from data
    collected in other studies.
  \item Trials 37, 38, and 39 do not exist. The numbers were accidentally
    skipped.
  \item Trials 9, 10, and 11 used a slightly different event definition where the
    calibration poses were not explicitly tagged by an event, yet the protocol
    was the identical to the following trials. The calibration pose will have
    to be determined manually.
  \item  Trials 6-15 have force measurements are affected by the treadmill
    vibration mode mentioned in the equipment section and the forces should be
    not be used. We include the trials because both the kinematic data is valid
    and trials 6-8 include lateral perturbations in addition to the
    longitudinal.
  \item During trials 9-15 we used wooden blocks to fix the treadmill to the
    concrete floor to eliminate the treadmill's low vibration mode
    (\~5\si{\hertz}). But these blocks seem to have corrupted the force plate
    measurements by imposing frictional stresses on the system. The force plate
    measurements should not be used from these trials, but the marker data is
    fine.
  \item Trials 6-8 use an early experimental protocol which divided the
    perturbation sections into three sections: longitudinal perturbations,
    lateral perturbations, and a combination of each. We then learned the
    treadmill had a low vibrational mode which significantly affects the force
    plate measurements, requiring us to eliminate the lateral perturbation
    motions. The force measurements during these trials are corrupted by this
    vibrational mode and should be used with caution or not at all.
  \item We did not record unloaded compensation trials for trials 9-15.
    Regardless, they would likely be useless due to the corruption from the
    wooden blocks.
  \item Trials 6-8 use a only the lower body marker set. The remaining trials
    are full body.
  \item The ankle joint torques computed from subject 9's data in trials 25-27
    are abnormal and should be used with caution or not at all. We were not
    able to locate the source of the error, but it is likely related to the
    force calibration.
\end{itemize}
